\Large Generative Dynamics of Supreme Court Citations

### 2018-04-30 Tex/paper.tex
updated holdout experiment results
revised network theory section

\title{\Large Inferential Analysis of the Supreme Court Citation Network}
%\author{}
 % Toggle % below to blind and unblind the manuscript
%  \author[1]{Christian Schmid \thanks{cxs5700@psu.edu}}
%   \author[2]{Ted Hsuan Yun Chen\thanks{thc126@psu.edu}}
% \author[2]{Bruce A. Desmarais\thanks{bdesmarais@psu.edu}}
% \author[1]{David R. Hunter \thanks{dhunter@stat.psu.edu}}
% \affil[2]{Department of Political Science, Pennsylvania State University}
 %\affil[1]{Department of Statistics, Pennsylvania State University}

@@

Majority opinions written by the United States Supreme Court exercise their authority and influence, in part, through their roles as precedents in future Supreme Court decisions and opinions. The findings regarding the extent and exact nature of the influences of precedent have been mixed, but the balance of the literature finds that past decisions exert some form of influence on the justices' decision making \citep{knight1996norm,gillman2001s,richards2002jurisprudential,hansford2006politics,bailey2008does,bailey2011constrained}. Despite a considerable body of research that focuses on the way in which relevant precedents shape decision making on the Court, relatively little work has focused on understanding which past opinions are cited by an opinion. Our focus in this paper is to provide what is, to our knowledge, the first comprehensive analysis of exactly which cases are cited by a case. We follow an emerging body of work on legal citations, and treat the system of citations as a network \citep[e.g., ][]{harris1982structural,caldeira1988legal,fowler2007network, fowler2008authority,bommarito2009law,lupu2012precedent,pelc2014politics}.

We are not the first to ask what predicts the citations in US Supreme Court Opinions. Indeed, a voluminous body of work has sought to explain how many times an opinion is cited \citep[e.g.,][]{cross2010determinants,benjamin2012standing}, when in a lifecycle an opinion is cited \citep[e.g.,][]{black2013citation,spriggs2001explaining}, and how many cases are cited by an opinion \citep[e.g.,][]{lupu2013strategic}---all focused on the US Supreme Court. One common feature of the research design in all of these studies is that the observations are at the case or case-time level. The outcome variables in these analyses are defined as measures of the number of citations to a case over a period of time, the number of citations to a case at a particular time, or a measurement on the cases cited by a case. None of these studies treats citation in its micro-level form---as a relationship between two opinions, the citing opinion and the cited opinion. We are aware of one prior study, \citet{clark2010locating}, in which a statistical model is used to analyze dyadic citations between cases. However, \citet{clark2010locating} use a dyadic latent variable model in order to estimate ideal points for Supreme Court Opinions, but does not use any explanatory variables to predict the formation of citation ties between opinions. We build upon this literature both methodologically and substantively. Methodologically, we develop a novel statistical framework for modeling directed dyadic citations. Second, we apply this methodology to a half-century of directed dyadic citations between US Supreme Court opinions.

There exist two broad reasons why empirical analyses of citations are best defined on the directed dyad level, not the case level. The first is that directed dyadic analyses can test both dyadic and case-level hypotheses. For example, case-level analyses can model whether opinions supported by a liberal majority coalition are more likely than those supported by a conservative majority coalition to be cited heavily in the future, but they cannot precisely model whether liberal cases will be cited more by liberal cases than by conservative cases. Thus, the first reason for analyzing citations at the dyadic level is to expand the set of hypotheses that can be represented in the model. The second reason for studying citations at the directed dyadic level is that, as articulated in the growing literature on legal citation networks, citations form complex networks in which a citation at one point in time may influence future citations. This phenomenon of complex dependence is very common in networks of many types, but processes specific to Supreme Court citations create interdependence in citations. For example, if opinion $i$ relies heavily on opinion $j$ as precedent, opinion $i$ is likely to discuss the legal basis for opinion $j$, and as a consequence cite some of the opinions cited by opinion $j$. Suppose opinion $k$ is cited by opinion $j$. Opinion $k$ is more likely to be cited by opinion $i$ because opinion $i$ relies heavily on $j$, and opinion $j$ cites $k$.  This is a special case of a very common process on networks referred to as ``triad closure''. Complex dependence is theoretically interesting on its own merits, but the effects of covariates cannot be reliably identified---either in terms of coefficient values or standard errors---without accounting for the interdependence inherent in networks \citep{desmaraisstatistical}.

In what follows we develop a theoretical case that citations on the US Supreme Court are characterized by forms of complex dependence that are common in networks. We then develop an extension of a model---the exponential random graph model---that can incorporate both exogenous covariates and complex forms of interdependence into a directed dyadic analysis of citations. Finally, we develop and estimate a specification of this model in an analysis of US Supreme Court citations between 1937 and 2001. We find robust support for the inherent complexity underpinning the formation of citation ties, and show that incorporating complex dependence into the model of citation formation significantly improves the model's predictive performance.\\[0.5cm]
\textit{Comment from Benjamin Kassov: I think the discussion in the introduction of the paper focusing on a "case-based" versus network based analysis is a bit short. 
I was left slightly confused by what a "case-based" analysis of citation actually is or what it means. 
I can see several different conceptions of this, some being more problematic than others in a methodological sense, 
although none of them are ideal for analyzing networks and all likely suffer some amount of OV bias.}

\section{Complex Interdependence in Supreme Court Citations}

When it comes to the development and testing of theory, the defining feature of networks is that the fundamental element under study---the relationship between two units (i.e., the citation from one opinion to another) is a piece of a complex web of relations. The formation (or lack thereof) of that relationship cannot be fully understood without considering how the relationship fits into the complex web. Analytical designs that account only for covariates in explaining tie formation are incomplete theoretically, and, as a consequence, are subject to a form of omitted variable bias \citep{cranmer2016critique}. Citations in legal opinions are unique in terms of the windows into network dependencies offered by the texts of the opinions. A number of common structural dependencies that are found in networks are likely to apply to citations in Supreme Court opinions. In this section we present these dependence forms, and document the mechanisms by which they arise through archetypal passages in example opinions.

\textbf{Transitivity:} In a network of directed relations (e.g., A cites B, but B doesn't cite A) transitivity refers to the tendency for A to send a tie to C if A sends a tie to B and B sends a tie to C \citep{holland1971transitivity}. In undirected networks, transitivity is simply the process by which friends of friends become friends (i.e., a friend of a friend is a friend). The term, ``transitive closure'' refers to a tie forming from A to C in response to extant ties from A to B and B to C. When writing opinions, Supreme Court justices present the legal bases for their rulings, which often involves discussing the most primary/relevant precedents underpinning these legal bases, but also the precedents and legal rules on which the primary precedents were based. This process of presenting several layers/levels of precedent in an opinion follows the structure of transitive closure exactly---opinion A cites opinion B as a primary precedent, and then cites opinion C because opinion B cites opinion C. The two examples presented below illustrate this process.

In the first example, a passage from Kansas v. Marsh (548 U.S. 163, 2006)---a case considering the constitutionality of a death sentence statute in Kansas. In this example, the case Stringer v. Black is cited by Kansas v. Marsh as a case that is quoted by Sochor v. Florida. The primary precedent under discussion in this passage of the opinion is Sochor v. Florida, but Stringer v. Black is cited as a result of its role in the Sochor v. Florida opinion.
\begin{quotation}
The statute thus addresses the risk of a morally unjustifiable death sentence, not by minimizing it as precedent unmistakably requires, but by guaranteeing that in equipoise cases the risk will be realized, by ``placing a `thumb [on] death's side of the scale,' '' Sochor v. Florida, 504 U. S. 527, 532 (1992) (quoting Stringer v. Black, 503 U. S. 222, 232 (1992); alteration in original).
\end{quotation} %https://supreme.justia.com/cases/federal/us/548/163/dissent2.html
The second example is a passage from Seminole Tribe of Fla. v. Florida, 
(517 U.S. 44 1996)---a case addressing the rights of groups and citizens to sue states in federal court. In this example, Pennsylvania v. Union Gas Co
(491 U.S. 1, 1989) is the primary precedent under discussion, and several cases are cited and discussed in terms of their roles as precedents in the Union Gas opinion. 
\begin{quotation}
Never before the decision in Union Gas had we suggested that the bounds of Article III could be expanded by Congress operating pursuant to any constitutional provision other than the Fourteenth Amendment. Indeed, it had seemed fundamental that Congress could not expand the jurisdiction of the federal courts beyond the bounds of Article III. Marbury v. Madison, 1 Cranch 137 (1803). The plurality's citation of prior decisions for support was based upon what we believe to be a misreading of precedent. See Union Gas, 491 U. S., at 40-41 (SCALIA, J., dissenting). The plurality claimed support for its decision from a case holding the unremarkable, and completely unrelated, proposition that the States may waive their sovereign immunity, see id., at 14-15 (citing Parden v. Terminal Railway of Ala. Docks Dept., 377 U. S. 184 (1964)), and cited as precedent propositions that had been merely assumed for the sake of argument in earlier cases, see 491 U. S., at 15 (citing Welch v. Texas Dept. of Highways and Public Transp., 483 U. S., at 475-476, and n. 5, and County of Oneida v. Oneida Indian Nation of N. Y., 470 U. S., at 252).'
\end{quotation}% https://supreme.justia.com/cases/federal/us/517/44/case.html

\noindent \textbf{Reciprocity:} Reciprocity (also referred to as mutuality) is the tendency for node B to send a tie to node A in response to or coordination with A sending a tie to B \citep{garlaschelli2004patterns}. It is typically not possible for reciprocated ties to form in legal opinions. Most citations reference past opinions that were issued before the citing opinion's case was even argued before the Court.  However, opinions written within the same Supreme Court term are often drafted in tandem, and can cite each other reciprocally.  Opinion A citing opinion B within the same term represents a signal that opinion B is relevant to the legal reasoning underpinning opinion A. Unlike the citations themselves, the applicability of legal rules or lines of reasoning across cases is not directed---if A is relevant to B, B is highly likely to be relevant to A. We expect opinions written within the same term to exhibit a high degree of reciprocity. Below we provide two example passages from opinions that illustrate the phenomenon of within-term reciprocity.

The first case in our example reciprocal dyad is a passage from Western Air Lines v. Criswell (472 U.S. 400, 1985)---a case considering mandatory retirement in the context of age discrimination laws. The second case in the dyad, Johnson v. Mayor of Baltimore (472 U.S. 353, 1985) is another case considering whether mandatory retirement violates the Age Discrimination in Employment Act. These cases addressed very similar legal questions, which increased the likelihood that they would inform each other, and the opinions were written within the same term, which made it possible for them to cite each other. 
\begin{quotation}
{\em From Western Air Lines:} On a more specific level, Western argues that flight engineers must meet the same stringent qualifications as pilots, and that it was therefore quite logical to extend to flight engineers the FAA's age 60 retirement rule for pilots. Although the FAA's rule for pilots, adopted for safety reasons, is relevant evidence in the airline's BFOQ defense, it is not to be accorded conclusive weight. Johnson v. Mayor and City Council of Baltimore, ante at 472 U. S. 370-371. The extent to which the rule is probative varies with the weight of the evidence supporting its safety rationale and "the congruity between the . . . occupations at issue." Ante at 472 U. S. 371. In this case, the evidence clearly established that the FAA, Western, and other airlines all recognized that the qualifications for a flight engineer were less rigorous than those required for a pilot.
\\~\\
{\em From Johnson:} The city, supported by several amici, argues for affirmance nonetheless. It asserts first that the federal civil service statute is not just a federal retirement provision unrelated to the ADEA, but in fact establishes age as a BFOQ for federal firefighters based on factors that properly go into that determination under the ADEA, see Western Air Lines, Inc. v. Criswell, post p. 472 U. S. 400. Second, the city asserts, a congressional finding that age is a BFOQ for a certain occupation is dispositive of that determination with respect to nonfederal employees in that occupation. 
\end{quotation} % https://supreme.justia.com/cases/federal/us/472/400/case.html \url{https://supreme.justia.com/cases/federal/us/472/353/case.html#370}

\noindent \textbf{Popularity} Popularity, also termed ``preferential attachment'' is the tendency for ties to be sent to nodes to which many ties have already been sent \citep{barabasi1999emergence}. Citations to an opinion signal both the Court's awareness of the legal reasoning of the case and the Court's evaluation that the opinion is an authoritative precedent. The more citations, the stronger this signal. Landmark cases, or those that establish new legal rules, are particularly authoritative and accrue citations from most future opinions that follow the respective line of reasoning. The passage below, from Oregon v. Mitchell (400 U.S. 112, 1970)---a case on the legality of state age restrictions on voting in federal elections---illustrates this popularity dynamic. In this opinion passage Baker v. Carr is cited in reference to its role as a landmark precedent, and noted for the number of other cases by which it has been followed and for which an authoritative opinion is referenced, and even discussed in terms of the number of other cases by which it was followed. 
\begin{quotation}
The first case in which this Court struck down a statute under the Equal Protection Clause of the Fourteenth Amendment was Strauder v. West Virginia, 100 U. S. 303, decided in the 1879 Term. [Footnote 2/1] In the 1961 Term, we squarely held that the manner of apportionment of members of a state legislature raised a justiciable question under the Equal Protection Clause, Baker v. Carr, 369 U. S. 186. That case was followed by numerous others, e.g.: that one person could not be given twice or 10 times the voting power of another person in a state-wide election merely because he lived in a rural area..."
\end{quotation} %https://supreme.justia.com/cases/federal/us/400/112/case.html

\noindent \textbf{Activity} Activity, or sender activation, is the tendency for ties sent to beget more ties sent \citep{howard2016understanding}. Structurally, this means that there are few nodes that send a moderate number of ties---the number of ties sent is either small or fairly large. In terms of Supreme Court opinions, for each opinion discussed, that discussion is likely to raise other tangential issues on which the justices will want to draw upon past opinions.  Furthermore, for each opinion that applies to and is cited by the current opinion, there is often a case/opinion that needs to be discussed in terms of why it does not apply. Justices often clarify not only those legal principles that apply, but often those that do not. The example passage that illustrates activity is from Kush v. Rutledge (460 U.S. 719, 1983)---a case addressing witness intimidation and due process protections. In the Kush opinion, Griffin v. Beckenridge (i.e., Griffin) is discussed at length in terms of its lack of applicability. It is common that opinions distinguish among a variety of legal rules in terms of their applicability to a case, which means that for each rule that is established in an applicable precedent, one or more need to be discussed in terms of why they do not apply.
\begin{quotation}
No allegations of racial or class-based invidiously discriminatory animus are required to establish a cause of action under the first part of 1985(2). The statutory provisions now codified at \S 1985 were originally enacted as \S 2 of the Civil Rights Act of 1871, and the substantive meaning of the 1871 Act has not been changed. The provisions relating to institutions and processes of the Federal Government (including the first part of \S 1985(2)) -- unlike those encompassing activity that is usually of primary state concern (including the second part of \S 1985(2) and the part of \P 1985(3) involved in Griffin, supra -- contain no language requiring that the conspirators act with intent to deprive their victims of the equal protection of the laws. Thus, the reasoning of Griffin is not applicable here, and, given the structure of \S 2 of the 1871 Act, it is clear''
\end{quotation} % https://supreme.justia.com/cases/federal/us/460/719/

\begin{itemize}
\item add to the discussion to which degree network statistics reflect causual relation
\end{itemize}

@@

@@

From this brief review we see two broad components of past methods for analyzing citation networks. First, node features, age in particular, are used to model the rate at which a document is cited. Second, theoretical models are used to account for prevalent graph structures. In the next section we describe a methodology that can be used to integrate node attributes and graph structures/motifs into a comprehensive model of US Supreme Court citations.

@@

@@

\begin{tabular}{rllll}
\hline \hline
& \multicolumn{2}{l}{Independent Model } & \multicolumn{2}{l}{Full Model} \\ \hline 
 & mean & range & mean & range \\ 
  \hline
precision & 0.5589 & (0.5508, 0.5757) & 0.8677 & (0.865, 0.8718) \\ 
  recall & 0.098 & (0.0943, 0.103) & 0.5943 & (0.5887, 0.6032) \\ 
  F1 score & 0.1667 & (0.1618, 0.1748) & 0.7054 & (0.7013, 0.7119) \\ 
   \hline \hline
\end{tabular}
\caption{The predictive performance of the directed dyadic models with linear trends, over ten 80/20 train/test splits.}
\label{tab:prediction_linear}

>>>>

\title{\Large Generative Dynamics of Supreme Court Citations}
%\author{}
 % Toggle % below to blind and unblind the manuscript
%  \author[1]{Christian Schmid \thanks{cxs5700@psu.edu}}
%   \author[2]{Ted Hsuan Yun Chen\thanks{thc126@psu.edu}}
% \author[2]{Bruce A. Desmarais\thanks{bdesmarais@psu.edu}}
% \author[1]{David R. Hunter \thanks{dhunter@stat.psu.edu}}
% \affil[2]{Department of Political Science, Pennsylvania State University}
 %\affil[1]{Department of Statistics, Pennsylvania State University}

@@

United States Supreme Court opinions exercise authority and influence, in part, through their roles as precedents affecting future jurisprudence in the US. The findings regarding the nature of the influences of precedent on the Supreme Court have been mixed, but the balance of the literature finds that past decisions exert some form of influence on the justices' decision making \citep{knight1996norm,gillman2001s,richards2002jurisprudential,hansford2006politics,bailey2008does,bailey2011constrained}. Despite a considerable body of research that focuses on the way in which relevant precedents shape decision making on the Court, relatively little work has focused on understanding which past opinions are cited by an opinion. Our focus in this paper is to provide what is, to our knowledge, the first comprehensive analysis of exactly which cases are cited in an opinion. We follow an emerging body of work on legal citations, and treat the system of citations as a network \citep[e.g., ][]{harris1982structural,caldeira1988legal,fowler2007network, fowler2008authority,bommarito2009law,lupu2012precedent,pelc2014politics}.

We are not the first to ask what predicts the citations in US Supreme Court Opinions. Indeed, a voluminous body of work has sought to explain how many times an opinion is cited \citep[e.g.,][]{cross2010determinants,benjamin2012standing}, when an opinion is cited \citep[e.g.,][]{black2013citation,spriggs2001explaining}, and how many cases are cited by an opinion \citep[e.g.,][]{lupu2013strategic}---all focused on the US Supreme Court. One common feature of the research design in all of these studies is that the observations are at the case or case-year level. The outcome variables in these analyses are defined as measures of the number of citations to a case over a period of time, the number of citations to a case at a particular time, or a measurement on the cases cited by a case. These are case-level studies in that, based on the unit of analysis, it is impossible to determine both the origin and target case of a citation that contributes to the dependent variable.

An alternative approach to case-level analysis of citations would be to model them in the directed dyadic form through which they arise. A case decided at time $t$ can cite (or not cite) each case decided previously, and in the US Supreme Court, each other case decided at time $t$. We are aware of one prior study, \citet{clark2010locating}, in which a statistical model is used to analyze directed dyadic citations between cases. However, \citet{clark2010locating} use a dyadic latent variable model in order to estimate ideal points for Supreme Court Opinions, but do not use this model to understand the relationships between explanatory variables and the formation of citation ties between opinions. We build upon the literature on citation analysis both methodologically and substantively. Methodologically, we develop a novel extension of a statistical model for networks, which we adapt to the network structural constraints of court citations. Second, we apply this methodology to a half-century of directed dyadic citations between US Supreme Court opinions.

There exist two broad benefits of conducting empirical analyses of citations at the directed dyad level. The first is that directed dyadic analyses can test both dyadic and case-level hypotheses. For example, case-level analyses can model whether opinions supported by a liberal majority coalition are more likely than those supported by a conservative majority coalition to be cited heavily in the future, but they cannot precisely model whether liberal cases will be cited more by liberal cases than by conservative cases. Thus, the first reason for analyzing citations at the dyadic level is to expand the set of hypotheses that can be represented in the model. The second reason for studying citations at the directed dyadic level is that, as articulated in the growing literature on legal citation networks, citations form complex networks in which a citation at one point in time may influence future citations. This phenomenon of complex dependence is very common in networks of many types, but processes specific to Supreme Court citations create interdependence in citations. For example, if opinion $i$ relies heavily on opinion $j$ as precedent, opinion $i$ is likely to discuss the legal basis for opinion $j$, and as a consequence cite some of the opinions cited by opinion $j$. Suppose opinion $k$ is cited by opinion $j$. Opinion $k$ is more likely to be cited by opinion $i$ because opinion $i$ relies heavily on $j$, and opinion $j$ cites $k$.  This is a special case of a very common process on networks referred to as ``triad closure''. Complex dependence is theoretically interesting on its own merits, but the effects of covariates cannot be reliably identified---either in terms of coefficient values or standard errors---without accounting for the interdependence inherent in networks \citep{desmaraisstatistical}.

In this paper we develop a theoretical case that citations on the US Supreme Court are characterized by forms of complex dependence that are common in networks. We then develop an extension of a model---the exponential random graph model (ERGM)---that can incorporate both exogenous covariates and complex forms of interdependence into a directed dyadic analysis of citations. Finally, we develop and estimate a specification of this model in an analysis of US Supreme Court citations between 1937 and 2015. We find robust support for the inherent complexity underpinning the formation of citation ties, and show that incorporating complex dependence into the model of citation formation significantly improves the model's predictive performance.

\section{Network Processes in Supreme Court Citations}

When it comes to the development and testing of theory, the defining feature of networks is that the fundamental element under study---the relationship between two units (i.e., the citation from one opinion to another) is a component of a complex system of relations. The formation (or lack thereof) of that relationship cannot be fully understood without considering how the relationship fits into the system. Analytical designs that account only for covariates in explaining tie formation are incomplete theoretically, and, as a consequence, are subject to a form of omitted variable bias \citep{cranmer2016critique}. Citations in legal opinions are unique in terms of the windows into network dependencies offered by the texts of the opinions. A number of common structural dependencies that are found in networks are likely to apply to citations in Supreme Court opinions. In this section we present these dependence forms, and document the mechanisms by which they arise through archetypal passages in example opinions.

We should note that we do not distinguish between positive and negative citations in this theoretical framework. The dynamics we outline are not specific to a particular type of citation. The only distinction we draw (in the empirical analysis) is the instance in which a case has been overruled. In the extreme instance of overruled precedents, we assume that (and test whether) a case is much less likely to be cited after it has been overruled.

The first network property that we theorize in the context of Supreme Court citations is transitivity. In a network of directed relations (e.g., A cites B, but B doesn't cite A) transitivity refers to the tendency for A to send a tie to C if A sends a tie to B and B sends a tie to C \citep{holland1971transitivity}. In undirected networks, transitivity is simply the process by which friends of friends become friends (i.e., a friend of a friend is a friend). The term, ``transitive closure'' refers to a tie forming from A to C in response to extant ties from A to B and B to C. When writing opinions, Supreme Court justices present the legal bases for their rulings, which often involves discussing the most primary/relevant precedents underpinning these legal bases, but also the precedents and legal rules on which the primary precedents were based. This process of presenting several layers/levels of precedent in an opinion follows the structure of transitive closure exactly---opinion A cites opinion B as a primary precedent, and then cites opinion C because opinion B cites opinion C. The two examples presented below illustrate this process.

\begin{figure}[htp]
\centering
\includegraphics[width = 0.95\textwidth,trim= 0cm 1cm 0cm 2cm,clip=true ]{images/citations_trans.pdf}
\caption{Illustration of transitive triangle connecting US Supreme Court opinions through citations.}
\label{fig:transitivity}
\end{figure}

In the first example, a passage from Kansas v. Marsh (548 U.S. 163, 2006)---a case considering the constitutionality of a death sentence statute in Kansas. In this example, the case Stringer v. Black is cited by Kansas v. Marsh as a case that is quoted by Sochor v. Florida. The primary precedent under discussion in this passage of the opinion is Sochor v. Florida, but Stringer v. Black is cited as a result of its role in the Sochor v. Florida opinion.
\begin{quotation}
The statute thus addresses the risk of a morally unjustifiable death sentence, not by minimizing it as precedent unmistakably requires, but by guaranteeing that in equipoise cases the risk will be realized, by ``placing a `thumb [on] death's side of the scale,' '' Sochor v. Florida, 504 U. S. 527, 532 (1992) (quoting Stringer v. Black, 503 U. S. 222, 232 (1992); alteration in original).
\end{quotation} %https://supreme.justia.com/cases/federal/us/548/163/dissent2.html
The second example, which we illustrate visually in Figure \ref{fig:transitivity} is a passage from Seminole Tribe of Fla. v. Florida, 
(517 U.S. 44 1996)---a case addressing the rights of groups and citizens to sue states in federal court. In this example, Pennsylvania v. Union Gas Co
(491 U.S. 1, 1989) is the primary precedent being critiqued, and several cases are cited and discussed in terms of their roles as precedents in the Union Gas opinion. We highlight one---Parden v. Terminal R. Co., which is cited and discussed in both the Seminole and Union Gas opinions. 
\begin{quotation}
Never before the decision in Union Gas had we suggested that the bounds of Article III could be expanded by Congress operating pursuant to any constitutional provision other than the Fourteenth Amendment. Indeed, it had seemed fundamental that Congress could not expand the jurisdiction of the federal courts beyond the bounds of Article III. Marbury v. Madison, 1 Cranch 137 (1803). The plurality's citation of prior decisions for support was based upon what we believe to be a misreading of precedent. See Union Gas, 491 U. S., at 40-41 (SCALIA, J., dissenting). The plurality claimed support for its decision from a case holding the unremarkable, and completely unrelated, proposition that the States may waive their sovereign immunity, see id., at 14-15 (citing Parden v. Terminal Railway of Ala. Docks Dept., 377 U. S. 184 (1964)), and cited as precedent propositions that had been merely assumed for the sake of argument in earlier cases, see 491 U. S., at 15 (citing Welch v. Texas Dept. of Highways and Public Transp., 483 U. S., at 475-476, and n. 5, and County of Oneida v. Oneida Indian Nation of N. Y., 470 U. S., at 252).'
\end{quotation}% https://supreme.justia.com/cases/federal/us/517/44/case.html

The second network property we consider in the context of Supreme Court citations is reciprocity. Reciprocity (also referred to as mutuality) is the tendency for node B to send a tie to node A in response to or coordination with A sending a tie to B \citep{garlaschelli2004patterns}. It is typically not possible for reciprocated ties to form in legal opinions. Most citations reference past opinions that were issued before the citing opinion's case was even argued before the Court.  However, opinions written within the same Supreme Court term are often drafted in tandem, and can cite each other reciprocally.  Opinion A citing opinion B within the same term represents a signal that opinion B is relevant to the legal reasoning underpinning opinion A. Unlike the citations themselves, the applicability of legal rules or lines of reasoning across cases is not directed---if A is relevant to B, B is highly likely to be relevant to A. We expect opinions written within the same term to exhibit a high degree of reciprocity. Below we provide two example passages from opinions that illustrate the phenomenon of within-term reciprocity.

\begin{figure}[htp]
\centering
\includegraphics[width = 0.95\textwidth,trim= 0cm 1cm 0cm 2cm,clip=true ]{images/citations_recip.pdf}
\caption{Illustration of a reciprocal tie between two US Supreme Court opinions.}
\label{fig:reciprocity}
\end{figure}

The first case in our example reciprocal dyad is a passage from Western Air Lines v. Criswell (472 U.S. 400, 1985)---a case considering mandatory retirement in the context of age discrimination laws. The second case in the dyad, Johnson v. Mayor of Baltimore (472 U.S. 353, 1985) is another case considering whether mandatory retirement violates the Age Discrimination in Employment Act. The mutual edge connecting these two cases is visualized in Figure \ref{fig:reciprocity}. These cases addressed very similar legal questions, which increased the likelihood that they would inform each other, and the opinions were written within the same term, which made it possible for them to cite each other. 
\begin{quotation}
{\em From Western Air Lines:} On a more specific level, Western argues that flight engineers must meet the same stringent qualifications as pilots, and that it was therefore quite logical to extend to flight engineers the FAA's age 60 retirement rule for pilots. Although the FAA's rule for pilots, adopted for safety reasons, is relevant evidence in the airline's BFOQ defense, it is not to be accorded conclusive weight. Johnson v. Mayor and City Council of Baltimore, ante at 472 U. S. 370-371. The extent to which the rule is probative varies with the weight of the evidence supporting its safety rationale and "the congruity between the . . . occupations at issue." Ante at 472 U. S. 371. In this case, the evidence clearly established that the FAA, Western, and other airlines all recognized that the qualifications for a flight engineer were less rigorous than those required for a pilot.
\\~\\
{\em From Johnson:} The city, supported by several amici, argues for affirmance nonetheless. It asserts first that the federal civil service statute is not just a federal retirement provision unrelated to the ADEA, but in fact establishes age as a BFOQ for federal firefighters based on factors that properly go into that determination under the ADEA, see Western Air Lines, Inc. v. Criswell, post p. 472 U. S. 400. Second, the city asserts, a congressional finding that age is a BFOQ for a certain occupation is dispositive of that determination with respect to nonfederal employees in that occupation. 
\end{quotation} % https://supreme.justia.com/cases/federal/us/472/400/case.html \url{https://supreme.justia.com/cases/federal/us/472/353/case.html#370}

The third network property we consider in the context of Supreme Court citations is popularity. Popularity, also termed ``preferential attachment'' is the tendency for ties to be sent to nodes to which many ties have already been sent \citep{barabasi1999emergence}. Citations to an opinion signal both the Court's awareness of the legal reasoning of the case and the Court's evaluation that the opinion is an authoritative precedent. The more citations, the stronger this signal. Landmark cases, or those that establish new legal rules, are particularly authoritative and accrue citations from most future opinions that follow the respective line of reasoning. The passage below, from Oregon v. Mitchell (400 U.S. 112, 1970)---a case on the legality of state age restrictions on voting in federal elections---illustrates this popularity dynamic. In this opinion passage Baker v. Carr is cited in reference to its role as a landmark precedent, and noted for the number of other cases by which it has been followed. and for which an authoritative opinion is referenced, and even discussed in terms of the number of other cases by which it was followed. The language in this passage suggests that the attention to Baker v. Carr in previous Court opinions is in part responsible for its authority in Oregon v. Mitchell. The citations to Baker v. Carr are visualized in Figrue \ref{fig:popularity}.
\begin{quotation}
The first case in which this Court struck down a statute under the Equal Protection Clause of the Fourteenth Amendment was Strauder v. West Virginia, 100 U. S. 303, decided in the 1879 Term. [Footnote 2/1] In the 1961 Term, we squarely held that the manner of apportionment of members of a state legislature raised a justiciable question under the Equal Protection Clause, Baker v. Carr, 369 U. S. 186. That case was followed by numerous others, e.g.: that one person could not be given twice or 10 times the voting power of another person in a state-wide election merely because he lived in a rural area..."
\end{quotation} %https://supreme.justia.com/cases/federal/us/400/112/case.html

\begin{figure}[htp]
\centering
\includegraphics[width = 0.95\textwidth,trim= 0cm 3cm 6cm 8cm,clip=true ]{images/citations_pop.pdf}
\caption{Illustration of ties sent to a landmark Supreme Court opinion via citations.}
\label{fig:popularity}
\end{figure}

The fourth network property we consider in the context of Supreme Court citations is activity. Activity, or sender activation, is the tendency for ties sent to beget more ties sent \citep{howard2016understanding}. Structurally, this means that there are few nodes that send a moderate number of ties---the number of ties sent is either small or fairly large. In terms of Supreme Court opinions, for each opinion discussed, that discussion is likely to raise other tangential issues on which the justices will want to draw upon past opinions.  Furthermore, for each opinion that applies to and is cited by the current opinion, there is often a case/opinion that needs to be discussed in terms of why it does not apply. Justices often clarify not only those legal principles that apply, but often those that do not. The example passage that illustrates activity is from Kush v. Rutledge (460 U.S. 719, 1983)---a case addressing witness intimidation and due process protections. In the Kush opinion, Griffin v. Beckenridge (i.e., Griffin) is discussed at length in terms of its lack of applicability. It is common that opinions distinguish among a variety of legal rules in terms of their applicability to a case, which means that for each rule that is established in an applicable precedent, one or more need to be discussed in terms of why they do not apply. We do not present a separate visualization related to activity, as it would look very similar to the visualization for popularity.
\begin{quotation}
No allegations of racial or class-based invidiously discriminatory animus are required to establish a cause of action under the first part of 1985(2). The statutory provisions now codified at \S 1985 were originally enacted as \S 2 of the Civil Rights Act of 1871, and the substantive meaning of the 1871 Act has not been changed. The provisions relating to institutions and processes of the Federal Government (including the first part of \S 1985(2)) -- unlike those encompassing activity that is usually of primary state concern (including the second part of \S 1985(2) and the part of \P 1985(3) involved in Griffin, supra -- contain no language requiring that the conspirators act with intent to deprive their victims of the equal protection of the laws. Thus, the reasoning of Griffin is not applicable here, and, given the structure of \S 2 of the 1871 Act, it is clear''
\end{quotation} % https://supreme.justia.com/cases/federal/us/460/719/

@@

\includegraphics[width = 0.95\textwidth,trim= 0cm 0cm 0cm 0cm,clip=true  ]{images/daggish.pdf}

@@

From this brief review we see two broad components of past methods for analyzing citation networks. First, node features, age in particular, are used to model the rate at which a document is cited. Second, theoretical models are used to account for prevalent graph structures. In the next section we describe a methodology that can be used to integrate case attributes and graph structures/motifs into a comprehensive model of US Supreme Court citations.

@@

Before we transition to the empirical analysis, we should note that there is nothing about the c-TERGM (or any ERGM family model, for that matter) that permits the identification of causal effects of one tie on the formation of another. For example, if we find a positive popularity effect, that could be driven by one of two (or some combination of both) dynamics. First, it could be the case that a large volume of citations to a case in past opinions draws the attention of the justices to the ``popular'' case. Second, it could be that we have omitted a case-level variable that is strongly associated with the number of citations accumulated by a case, and the number of citations in past cases is operating as a proxy for this unmeasured case characteristic. We cannot statistically differentiate between these two dynamics. However, in both of these conditions, and analogous conditions related to other dependence properties, citations (or non-citations) to a case are not independent conditional on the other variables, and the c-TERGM can be specified to account for the dependence in the data.

@@

\begin{tabular}{rllll}
\hline \hline
& \multicolumn{2}{l}{Independent Model } & \multicolumn{2}{l}{Full Model} \\ \hline 
 & mean & range & mean & range \\ 
  \hline
precision & 0.6007 & (0.5934, 0.6074) & 0.8666 & (0.8647, 0.8701) \\ 
  recall & 0.1522 & (0.1495, 0.1549) & 0.5969 & (0.5945, 0.6019) \\ 
  F1 score & 0.2429 & (0.2389, 0.2462) & 0.7069 & (0.7048, 0.7107) \\ 
   \hline \hline
\end{tabular}
\caption{The predictive performance of the directed dyadic models with linear trends, over ten 80/20 train/test splits.}
\label{tab:prediction_linear}
